Linking watershed nutrient loading to estuary water quality with semiparametric models

### 2023-06-23 Schramm-2023-06-PeerJ/Schramm-2023-06-PeerJ.Rmd
peerj submission files

---
title: Linking watershed nutrient loading to estuary water quality with semiparametric models
preprint: false
author: 
  - name: Michael Schramm
    affiliation: 1
    corresponding: true
    email: michael.schramm@ag.tamu.edu
affiliation:
  - code: 1
    address: Texas Water Resources Institute, Texas A\&M AgriLife Research, College Station, Texas, United States
abstract: >
  Evaluating estuary water quality responses to reductions (or increases) in nutrient loading attributed to on the ground management actions can be challenging due to the strong influence of environmental drivers on nutrient loads and non-linear relationships. This study applied semiparametric approaches to calculate watershed nutrient loads and assess responses in estuary water quality to seasonally-adjusted freshwater inflow and flow-adjusted nutrient loads in Lavaca Bay, Texas. Lavaca Bay is a secondary embayment on the Texas coast displaying early potential for eutrophication and water quality degredation. Use of flow-adjusted nutrient loads allowed the study to evaluate the response in water quality to changes in nutrient loads driven by non-natural variation. Cross-validation indicated that, despite data constraints, semiparametric models performed well at nutrient load prediction. Based on these models, delivered annual nutrient loads varied substantially from year to year. In contrast, minimal changes in flow-normalized loads indicate that nutrient loadings were driven by natural variation in precipitation and runoff as opposed to changes in management of nonpoint sources. Models indicated no evidence of long-term changes in dissolved oxygen or chlorophyll-*a* within Lavaca Bay. However, site specific long-term increases in both organic and inorganic nitrogen are causes of concern as potential causes of future eutrophication. Further analysis found freshwater inflow was a strong driver of nutrient and chlorophyll-*a* concentrations within Lavaca Bay but there was no evidence that changes in watershed nutrient loading explained additional variation in dissolved oxygen or chlorophyll-*a* concentrations. In addition to providing a baseline assessment of watershed nutrient loading and water quality responses in the Lavaca Bay watershed, this study provides methodological support for the use of semiparametric methods in load regression models and estuary assessments.
bibliography: bibliography.bib
output:
  rticles::peerj_article:
    citation_package: biblatex
biblio-style: apa
header-includes:
    - \usepackage{csquotes}

@@

## this sets our default code chunk options
knitr::opts_chunk$set(echo = FALSE,
                      ## figure chunk options
                      dev = "ragg_png",
                      dpi = 300,
                      fig.align = "center",
                      out.width = "100%")

@@

Similar to many coastal areas globally, the coastal watersheds along the Texas Gulf coast 
are facing pressures from increasing population, increases in point source 
and non-point source pollution and alterations to freshwater flows that degrade
water quality in downstream estuaries [@bricker_effects_2008; @kennicuttWaterQualityGulf2017; @bugica_water_2020]. 
Despite these escalating pressures, national scale assessments have classified 
coastal estuaries in Texas as moderate or low risk for eutrophic conditions 
[@bricker_effects_2008]. However, a suite of recent studies indicates that 
estuary water quality dynamics in both agriculturally dominated and urban 
watersheds within Texas are expressing conditions that are increasingly 
conducive to algal blooms and eutrophication [@wetzWaterQualityDynamics2016; 
@wetz_exceptionally_2017; @bugica_water_2020; @chinPhytoplanktonBiomassCommunity2022]. 
With identification of several localized areas of estuary water quality concern along 
the Texas coast [@bugica_water_2020], localized studies are being prioritized 
to better inform management actions.

This goal of this project is to assess watershed nutrient loading and the 
resulting water quality responses in Lavaca Bay, Texas. Lavaca Bay is a secondary bay in the larger
Matagorda Bay system located roughly halfway between Houston, Texas and Corpus
Christi, Texas. Lavaca Bay faces substantial challenges associated with
legacy contamination but general water quality parameters such as dissolved oxygen 
(DO), nutrients, and biological parameters have been well within state
water quality standards. Despite largely meeting statewide water quality standards,
there have been concerning declines in abundance, biomass, and diversity of 
benthic fauna in Lavaca Bay [@beserespollackLongtermTrendsResponse2011]. These declines
are partially attributed to reductions in freshwater inflow and changes in estuary salinity
and are indicative of an already stressed system 
[@beserespollackLongtermTrendsResponse2011; @palmerImpactsDroughtsLow2015;
@montagnaAssessmentRelationshipFreshwater2020]. More recently, significant
linear increases in total phosphorus (TP), orthophosphate, 
total Kjeldahl nitrogen (TKN), and chlorophyll-*a* concentrations were identified
at monitoring sites within Lavaca Bay [@bugica_water_2020]. Although 
long-term changes in DO concentrations have not been identified, the trends in 
nutrient concentrations are concerning due to the role of nitrogen as a 
limiting factor for primary production in many Texas estuaries 
[@gardnerNitrogenFixationDissimilatory2006; @houTransformationFateNitrate2012;
@doradoUnderstandingInteractionsFreshwater2015; 
@paudelRelationshipSuspendedSolids2019; @wetz_exceptionally_2017] and the 
ramifications that changes in nitrogen loadings could have for 
productivity and eutrophication in Lavaca Bay.

There are ongoing efforts between local, state, and federal agencies to 
address water quality impairments in the freshwater portions of the 
Lavaca Bay watershed [@jainTechnicalSupportDocument2021; 
@schrammLavacaRiverWatershed2018; @bertholdDirectMailingEducation2021]. 
However, at a statewide scale, these approaches have shown limited success and 
emphasize a need for improved efforts at assessing and linking management 
actions with downstream water quality to identify and replicate effective 
management actions across the state [@schrammTotalMaximumDaily2022]. The 
identification and communication of changes and trends in water quality is 
complicated by the fact that trends are often non-linear and confounded by 
precipitation and runoff that hinder traditional analysis 
[@wazniakLinkingWaterQuality2007; @lloydMethodsDetectingChange2014]. 
The development and application of flexible statistical methods such as Weighted
Regressions on Time, Discharge and Season [WRTDS, @hirschWeightedRegressionsTime2010] 
and Generalized Additive Models [GAMs, @woodFastStableRestricted2011] have 
provided effective tools for researchers to quantify and
communicate non-linear changes in river and estuary pollutant loadings.

WRTDS calculates a time series of in-stream concentrations or loads 
(daily, monthly, or annually) and flow-normalized estimates of concentrations and loads
using locally weighted regresssion for unique combinations of time, discharge, and season. 
WRTDS has been widely used to assess and identify trends in riverine nutrients 
[@oelsner_recent_2019; @stackpooleLongTermMississippi2021],
chlorides [@stetsIncreasingChlorideRivers2018], and other pollutants of concern 
[@shodaWaterqualityTrendsRivers2019]. WRTDS has also been succesfully adapted to 
assess trends in estuarine water quality concentrations [@beckFourDecadesWater2018].

While WRTDS is a statistical approach developed specifically for water quality applications, 
GAMs are a broadly applicable statistical method. GAMs are a semiparametric 
extension of generalized linear models where the linear predictor is represented
as the sum of multiple unknown smooth functions and parametric linear predictors
[@woodFastStableRestricted2011]. Although the underlying parameter estimation 
procedure of GAMs are substantially different than WRTDS, both the functional 
form and results have been demonstrated to be similar when assessing nutrient
concentration trends [@beckNumericalQualitativeContrasts2017]. 
Water quality applications of GAMs have included river and catchment 
nutrient concentration and load estimation
[@wangLoadEstimationUncertainties2011;@kroonRiverLoadsSuspended2012;
@kuhnertQuantifyingTotalSuspended2012;@robson_prediction_2015-1;
@hagemannEstimatingNutrientOrganic2016;@mcdowell_implications_2021;
@biagi_novel_2022], and assessment of temporal trends of nutrients
[@beckNumericalQualitativeContrasts2017; @murphyGeneralizedAdditiveModel2019],
phytoplankton [@bergbuschUnexpectedShiftPhytoplankton2021], 
and cyanobacteria [@hayesEffectsLakeWarming2020]. 
Recently GAMs have also been used to link water quality responses in 
receiving water bodies to changes in nonpoint source nutrient inputs 
[@murphyNutrientImprovementsChesapeake2022]. For a substantial discussion on the 
differences (and similarities) between GAMs and WRTDS for water quality 
applications reders are referred to @beckNumericalQualitativeContrasts2017.

@@

knitr::include_graphics("../fig1.png")

@@

Water quality data for the two freshwater and three estuary locations were 
obtained from the Texas Commission on Environmental Quality (TCEQ) Surface Water 
Quality Monitoring Information System. Data submitted through the system 
are required to be collected under Quality Assurance Project Plans and lab 
method procedures outlined by the TCEQ's procedures manual to ensure
consistent collection and laboratory methods are applied between samples 
collected by different entities and under different projects. 
All sites had varying lengths of and availability of data.
For freshwater locations, TP from January 2000 through December 2020 and 
nitrate-nitrogen (NO~3~) data from January 2005 through December 2020 were 
downloaded (Table\ \ref{tab:table1}). Less than 5-years of total
nitrogen and TKN concentration data were available at the freshwater sites and
deemed insufficient to develop load estimation models 
[@horowitzEvaluationSedimentRating2003; @snelderEstimationCatchmentNutrient2017]. 
The three estuary sites included an upper Lavaca Bay site near the outlet of the
Lavaca River system (TCEQ-13563), a mid-Lavaca Bay site (TCEQ-13383), and the 
lower Lavaca Bay site near the mouth of the Bay (TCEQ-13384). 
For estuary locations, we obtained data for TP, Nitrite+Nitrate (NO~*X*~), TKN, 
chlorophyll-*a*, and DO concentrations from January 2005 through December 2020 
(Table\ \ref{tab:table2}).

@@

\begin{align}
g(\mu) &= \alpha + f_1(ddate) + f_2(yday) + f_3(log1p(Q)) + f_4(ma) + f_5(fa), \nonumber \\
y &\sim \mathcal{N}(\mu,\,\sigma^{2})
\label{eq1}
\end{align}

where $\mu$ is the conditional expected NO~3~ or TP concentration, $g()$ is the 
log-link, $\alpha$ is the intercept, $f_n()$ are smoothing functions. $y$ is the 
response variable (NO~3~ or TP concentration) modeled as normally distributed 
with mean $\mu$ and standard deviation $\sigma$. *ddate* is the date converted to 
decimal notation, *yday* is numeric day of year (1-366), and *log1p(Q)* is the 
natural log of mean daily streamflow plus 1.

@@

Daily loads were estimated as the predicted concentration multiplied by the 
daily streamflow. For the Navidad River (USGS-08164525) site, daily loads at the 
dam were calculated 
from the discrete daily concentration at the discharge point of the lake and 
corresponding reported daily discharge from the dam.
Flow-normalized loads were estimated similar to WRTDS by setting flow-based 
covariates on each day of the year equal to each of the historical values for
that day of the year over the study period [@hirschWeightedRegressionsTime2010]. The 
flow-normalized estimate was calculated as the mean of all the predictions for 
each day considering all possible flow values.
Standard deviations and 90% credible intervals were obtained by 
drawing samples from the multivariate normal posterior distribution of the 
fitted GAM [@woodConfidenceIntervalsGeneralized2006; 
@marraCoveragePropertiesConfidence2012; @mcdowell_implications_2021]. 
GAM performance was evaluated 
using repeated 5-fold cross validation  [@burmanComparativeStudyOrdinary1989] 
and average Nash-Sutcliffe Efficiency (NSE), Pearson sample correlation (*r*) 
and percent bias (PBIAS) metrics across folds were calculated for each model.

@@

Because streamflow and nutrient loads are tightly correlated, freshwater inflow
can mask signals due to chnages in nutrient loads alone. Following the 
methodology implemented by @murphyNutrientImprovementsChesapeake2022,
both freshwater inflow and nutrient loads were prepossessed to account for season
and streamflow respectively. Raw freshwater inflow values were replaced by 
seasonally adjusted log transformed inflow obtained from the residuals of a GAM model 
fit between season(day of year) and log transformed daily freshwater inflow. Raw nutrient
loads were replaced with flow-adjusted values obtained from the residuals of a
GAM model relating log transformed NO~3~ or TP loads to log transformed daily
inflow. Response residuals from the respective GAM models were 
used as *Q* and *Load* in Eq.\ \ref{eq6} and Eq.\ \ref{eq7}.

@@

Predictive performance of nutrient loads ranged from "satisfactory" to "very good"
based on standardized evaluation metrics of NSE, *r*, and PBIAS 
[@moriasiHydrologicWaterQuality2015] calculated using 5-fold cross validation.
Median goodness-of-fit metrics for NO~3~ models in the Lavaca 
River were 0.34 NSE, 0.70 *r*, and 2.00 PBIAS. Navidad River NO~3~ models 
appeared to perform slightly better with 0.48 NSE and 0.87 *r* but with higher 
bias at 10.90 PBIAS. Generally, TP models performed better than NO~3~ models. 
Median goodness-of-fit metrics for TP in the Lavaca River were 0.81 NSE, 
0.93 *r*, and -7.20 PBIAS. Navidad River TP models has similar performance 
with 0.91 NSE, 0.99 *r*, and -3.30 PBIAS. Density plots of
metrics show similar distribution of values between sites for the 
same parameter, with the exception *r* values for NO~3~ loads where 
Lavaca River had a much larger variance in values compared to the Navidad River
(Fig.\ \ref{fig:fig2}). TP GAMS had higher average NSE and *r* values and 
lower variance in metric values compared to NO~3~.

```{r fig2, warning=FALSE, message=FALSE, fig.width=6.85, fig.asp=0.65, fig.cap="Density plots of goodness-of-fit metrics (NSE, \\textit{r}, and PBIAS) from repeated 5-fold cross validation between predicted nutrient loads from GAM models and measured nutrient loads. Color indicates the tail probability calculated from the empirical cumulative distribution of the goodness-of-fit metrics."}

@@

```{r fig3, error=FALSE, warning=FALSE, message=FALSE, fig.width=6.85, fig.asp=0.7, fig.cap="Aggregated estimated annual and flow-normalized annual NO\\textsubscript{3} and TP loads for USGS-08164000 and USGS-08164525."}

@@

Aggregated across both sites, the mean 
annual NO~3~ load 2005 through 2020 was `r mean_no3` kg (`r mean_no3_low` kg - 
`r mean_no3_upper` kg, 90% CI). Annual NO~3~ loads ranged from `r min_no3` kg 
in `r min_no3_year` to `r max_no3` kg in `r max_no3_year`. Total annual TP loads 
ranged from `r min_tp` kg in `r min_tp_year` to 
`r max_tp` kg  in `r max_tp_year`. Mean annual TP loading from 2005 through 2020
was `r mean_tp` kg (`r mean_tp_low` kg - `r mean_tp_upper` kg, 90% CI).
On average, the Navidad River accounted for `r no3_mean_prop` of NO~3~ loads and 
`r tp_mean_prop` of TP loads from 2005 through 2020. However, during periods of extreme 
drought the Lavaca River became the primary source of 
nutrient loading in the watershed with the Navidad River only accounting for 
`r min_no3_prop` and `r min_tp_prop` of NO~3~ and TP loads in 2011 
(Fig.\ \ref{fig:fig4}).

```{r fig4, error=FALSE, message=FALSE, warning=FALSE, fig.width=3.3, fig.asp=2, fig.cap="Comparison of delivered annual loads and annual discharge at the Lavaca (USGS-08164000) and Navidad (USGS-08164525) Rivers.", out.width="0.5\\linewidth"}

@@

```{r fig5, message=FALSE, warning=FALSE, fig.width=6.85, fig.asp=0.7, fig.cap="Fitted splines (shaded regions indicate 90\\% confidence intervals) from the temporal estuary GAM display the marginal smoothed effect of date on TP (A), NO\\textit{\\textsubscript{X}} (B), chlorophyll-\\textit{a} (C), TKN (D), and DO (E) concentrations at each site in Lavaca Bay."}

@@

```{r fig6, message=FALSE, warning=FALSE, fig.width=6.85, fig.asp=0.7, fig.cap="Fitted splines from estuary GAMs display the marginal smoothed effect of freshwater inflow (controlled for season) on TP (A), NO\\textit{\\textsubscript{X}} (B), chlorophyll-\\textit{a} (C), TKN (D), and DO (E) concentrations at each site in Lavaca Bay."}

@@

```{r fig7, message=FALSE, warning=FALSE, fig.width=6.85, fig.asp=0.4, fig.cap="Fitted splines from the nutrient loading GAMs display the marginal smoothed effect of TP and NO\\textsubscript{3} loads on TP (A) and NO\\textsubscript{\\textit{X}} (B) concentrations at each site in Lavaca Bay."}

@@

knitr::include_graphics("../fig8.png")

@@

The non-linear estuary water quality trends identified in the current
study differed slightly from previously identified trends [@bugica_water_2020]. 
This is not unexpected due to the different time periods, different methodology, and
generally small slopes identified for most of the significant water quality
parameters in prior work. Both DO and cholorophyll-*a* concentrations at all three
Lavaca Bay sites were stable from 2005 through 2020. This is a positive outcome 
in comparison to other Texas estuaries that are facing larger demands for 
freshwater diversions, higher population growth, and more intense agricultural 
production which have resulted in more direct signs of eutrophication
[@wetzWaterQualityDynamics2016; @bugica_water_2020]. Despite the stability of 
DO and cholorophyll-*a*, there are concerning site specific increases in 
NO~*X*~ and TKN concentration over the same time period. These trends are espcially
concerning due to the nitrogen limitation identified in many Texas 
estuaries [@gardnerNitrogenFixationDissimilatory2006; 
@houTransformationFateNitrate2012; @doradoUnderstandingInteractionsFreshwater2015; 
@paudelRelationshipSuspendedSolids2019; @wetz_exceptionally_2017] 
and the relatively low ambient concentrations observed in Lavaca Bay.

>>>>

---
title: Linking watershed nutrient loading to estuary water quality with semiparametric models
preprint: false
author: 
  - name: Michael Schramm
    affiliation: 1
    corresponding: true
    email: michael.schramm@ag.tamu.edu
affiliation:
  - code: 1
    address: Texas Water Resources Institute, Texas A\&M AgriLife Research, College Station, Texas, United States
abstract: >
  Evaluating estuary water quality responses to reductions (or increases) in nutrient loading attributed to on the ground management actions can be challenging due to the strong influence of environmental drivers on nutrient loads and non-linear relationships. This study applied semiparametric approaches to calculate watershed nutrient loads and assess responses in estuary water quality to seasonally-adjusted freshwater inflow and flow-adjusted nutrient loads in Lavaca Bay, Texas. Lavaca Bay is a secondary embayment on the Texas coast displaying early potential for eutrophication and water quality degradation. Use of flow-adjusted nutrient loads allowed the study to evaluate the response in water quality to changes in nutrient loads driven by non-natural variation. Cross-validation indicated that, despite data constraints, semiparametric models performed well at nutrient load prediction. Based on these models, delivered annual nutrient loads varied substantially from year to year. In contrast, minimal changes in flow-normalized loads indicate that nutrient loadings were driven by natural variation in precipitation and runoff as opposed to changes in management of nonpoint sources. Models indicated no evidence of long-term changes in dissolved oxygen or chlorophyll-*a* within Lavaca Bay. However, site specific long-term increases in both organic and inorganic nitrogen are concerning for their potential to fuel eutrophication. Further analysis found freshwater inflow had strong influences on nutrient and chlorophyll-*a* concentrations but there was no evidence that changes in watershed nutrient loading explained additional variation in dissolved oxygen or chlorophyll-*a* concentrations. In addition to providing a baseline assessment of watershed nutrient loading and water quality responses in the Lavaca Bay watershed, this study provides methodological support for the use of semiparametric models in load regression models and estuary assessments.
bibliography: bibliography.bib
output:
  rticles::peerj_article:
    citation_package: biblatex
biblio-style: apa
header-includes:
    - \usepackage{csquotes}

@@

## this sets our default code chunk options
knitr::opts_chunk$set(echo = FALSE,
                      ## figure chunk options
                      dev = "ragg_png",
                      dpi = 300,
                      fig.align = "center",
                      out.width = "100%",
                      fig.process = 
                        function(filename) {
                          new_filename <- stringr::str_remove(string = filename,
                                                              pattern = "-1")
                          fs::file_move(path = filename, new_path = new_filename)
                          ifelse(fs::file_exists(new_filename), new_filename, filename)
                        }
                      )

@@

Similar to many coastal areas globally, the coastal watersheds along the Texas 
Gulf coast are facing pressures from growing populations, increases in point source 
and non-point source pollution and alterations to freshwater flows that degrade
water quality in downstream estuaries [@bricker_effects_2008; 
@kennicuttWaterQualityGulf2017; @bugica_water_2020]. 
Despite these escalating pressures, national scale assessments have classified 
coastal estuaries in Texas as moderate or low risk for eutrophic conditions 
[@bricker_effects_2008]. However, a suite of recent studies indicates that 
estuary water quality dynamics in both agricultural and urban dominated
watersheds within Texas are expressing conditions that are increasingly 
conducive to algal blooms and eutrophication [@wetzWaterQualityDynamics2016; 
@wetz_exceptionally_2017; @bugica_water_2020; @chinPhytoplanktonBiomassCommunity2022]. 
With identification of several localized areas of estuary water quality concern 
along the Texas coast [@bugica_water_2020], localized studies are being prioritized 
to better inform management actions.

The goal of this project is to assess watershed nutrient loading and the 
resulting water quality responses in Lavaca Bay, Texas. Lavaca Bay is a secondary bay in the larger
Matagorda Bay system located roughly halfway between Houston, Texas and Corpus
Christi, Texas. Lavaca Bay faces substantial challenges associated with
legacy contamination but general water quality parameters such as dissolved oxygen 
(DO), nutrients, and biological parameters have been well within state
water quality standards. Despite largely meeting state designated water quality 
thresholds, there have been concerning declines in abundance, biomass, and 
diversity of benthic fauna in Lavaca Bay [@beserespollackLongtermTrendsResponse2011]. 
These declines are partially attributed to reductions in freshwater inflow and 
changes in estuary salinity and are indicative of an already stressed system 
[@beserespollackLongtermTrendsResponse2011; @palmerImpactsDroughtsLow2015;
@montagnaAssessmentRelationshipFreshwater2020]. More recently, significant
linear increases in total phosphorus (TP), orthophosphate, 
total Kjeldahl nitrogen (TKN), and chlorophyll-*a* concentrations were identified
at monitoring sites within Lavaca Bay [@bugica_water_2020]. These trends in 
nutrient concentrations are concerning due to the role of nitrogen as a 
limiting factor for primary production in many Texas estuaries 
[@gardnerNitrogenFixationDissimilatory2006; @houTransformationFateNitrate2012;
@doradoUnderstandingInteractionsFreshwater2015; 
@paudelRelationshipSuspendedSolids2019; @wetz_exceptionally_2017] and the 
ramifications that changes in nitrogen loadings could have for 
productivity and eutrophication in Lavaca Bay.

There are ongoing efforts between local, state, and federal agencies to 
address water quality impairments in the freshwater portions of the 
Lavaca Bay watershed [@jainTechnicalSupportDocument2021; 
@schrammLavacaRiverWatershed2018; @bertholdDirectMailingEducation2021]. 
However, on a statewide scale, these approaches have shown limited success and 
emphasize a need for improved methods to assess and link management 
actions with downstream water quality [@schrammTotalMaximumDaily2022].  
Such methods could help identify and replicate effective water quality 
management actions across the state . The identification and communication of 
changes and trends in water quality is 
complicated by the fact that trends are often non-linear and confounded by 
precipitation and runoff that hinder traditional analysis 
[@wazniakLinkingWaterQuality2007; @lloydMethodsDetectingChange2014]. 
The development and application of flexible statistical methods such as Weighted
Regressions on Time, Discharge and Season [WRTDS, @hirschWeightedRegressionsTime2010] 
and Generalized Additive Models [GAMs, @woodFastStableRestricted2011] have 
provided effective tools for researchers to quantify and
communicate non-linear changes in river and estuary pollutant loadings.

WRTDS calculates a time series of in-stream concentrations or loads 
(daily, monthly, or annually) and flow-normalized estimates of concentrations and loads
using locally weighted regression for unique combinations of time, discharge, and season. 
WRTDS has been widely used to assess and identify trends in riverine nutrients 
[@oelsner_recent_2019; @stackpooleLongTermMississippi2021],
chlorides [@stetsIncreasingChlorideRivers2018], and other pollutants of concern 
[@shodaWaterqualityTrendsRivers2019]. WRTDS has also been successfully adapted to 
assess trends in estuary water quality concentrations [@beckFourDecadesWater2018].

While WRTDS is a statistical approach developed specifically for water quality applications, 
GAMs are a broadly applicable statistical method. GAMs are a semiparametric 
extension of generalized linear models where the linear predictor is represented
as the sum of multiple unknown smooth functions and parametric linear predictors
[@woodFastStableRestricted2011]. Although the underlying parameter estimation 
procedure of GAMs are substantially different than WRTDS, both the functional 
form and results have been demonstrated to be similar when assessing nutrient
concentration trends [@beckNumericalQualitativeContrasts2017]. 
Water quality applications of GAMs have included river and catchment 
nutrient concentration and load estimation
[@wangLoadEstimationUncertainties2011;@kroonRiverLoadsSuspended2012;
@kuhnertQuantifyingTotalSuspended2012;@robson_prediction_2015-1;
@hagemannEstimatingNutrientOrganic2016;@mcdowell_implications_2021;
@biagi_novel_2022], and assessment of temporal trends of nutrients
[@beckNumericalQualitativeContrasts2017; @murphyGeneralizedAdditiveModel2019],
phytoplankton [@bergbuschUnexpectedShiftPhytoplankton2021], 
and cyanobacteria [@hayesEffectsLakeWarming2020]. 
Recently GAMs have also been used to link water quality responses in 
receiving water bodies to changes in nonpoint source nutrient inputs 
[@murphyNutrientImprovementsChesapeake2022]. For a substantial discussion on the 
differences (and similarities) between GAMs and WRTDS for water quality 
applications readers are referred to @beckNumericalQualitativeContrasts2017.

@@

knitr::include_graphics("Schramm-2023-06-PeerJ_files/figure-latex/fig1.png")

@@

Water quality data for the two freshwater and three estuary locations were 
obtained from the Texas Commission on Environmental Quality (TCEQ) Surface Water 
Quality Monitoring Information System. Data submitted through the system 
are required to be collected under Quality Assurance Project Plans and lab 
method procedures outlined by the TCEQ's procedures manual. These operating 
procedures ensure consistent collection and laboratory methods are applied 
between samples collected by different entities and under different projects. 
All sites had varying time periods and availability of water quality data.
For freshwater locations, TP from January 2000 through December 2020 and 
nitrate-nitrogen (NO~3~) data from January 2005 through December 2020 were 
downloaded (Table\ \ref{tab:table1}). Less than 5-years of total
nitrogen and TKN concentration data were available at the freshwater sites and
deemed insufficient to develop load estimation models 
[@horowitzEvaluationSedimentRating2003; @snelderEstimationCatchmentNutrient2017]. 
The three estuary sites included an upper Lavaca Bay site near the outlet of the
Lavaca River system (TCEQ-13563), a mid-Lavaca Bay site (TCEQ-13383), and the 
lower Lavaca Bay site near the mouth of the Bay (TCEQ-13384). 
For estuary locations, we obtained data for TP, Nitrite+Nitrate (NO~*X*~), TKN, 
chlorophyll-*a*, and DO concentrations from January 2005 through December 2020 
(Table\ \ref{tab:table2}).

@@

\begin{align}
g(\mu) &= \alpha + f_1(ddate) + f_2(yday) + f_3(log(1+Q)) + f_4(ma) + f_5(fa), \nonumber \\
y &\sim \mathcal{N}(\mu,\,\sigma^{2})
\label{eq1}
\end{align}

where $\mu$ is the conditional expected NO~3~ or TP concentration, $g()$ is the 
log-link, $\alpha$ is the intercept, $f_n()$ are smoothing functions. $y$ is the 
response variable (NO~3~ or TP concentration) modeled as normally distributed 
with mean $\mu$ and standard deviation $\sigma$. *ddate* is the date converted to 
decimal notation, *yday* is numeric day of year (1-366), and *Q* is the 
mean daily streamflow.

@@

Daily loads were estimated as the predicted concentration multiplied by the 
daily streamflow. For the Navidad River (USGS-08164525) site, daily loads at the 
dam were calculated 
from the discrete daily concentration at the discharge point of the lake and 
corresponding reported daily discharge from the dam.
Flow-normalized loads were estimated similar to WRTDS by setting flow-based 
covariates on each day of the year equal to each of the historical values for
that day of the year over the study period [@hirschWeightedRegressionsTime2010]. The 
flow-normalized estimate was calculated as the mean of all the predictions for 
each day considering all possible flow values.
Standard deviations and 90% credible intervals were obtained by 
drawing samples from the multivariate normal posterior distribution of the 
fitted GAM [@woodConfidenceIntervalsGeneralized2006; 
@marraCoveragePropertiesConfidence2012; @mcdowell_implications_2021]. 
GAM performance was evaluated 
with repeated 5-fold cross validation  [@burmanComparativeStudyOrdinary1989] 
using average Nash-Sutcliffe Efficiency (NSE), Pearson sample correlation (*r*) 
and percent bias (PBIAS) metrics.

@@

Because streamflow and nutrient loads are tightly correlated, freshwater inflow
can mask signals due to changes in nutrient loads alone. Following the 
methodology implemented by @murphyNutrientImprovementsChesapeake2022,
both freshwater inflow and nutrient loads were prepossessed to account for season
and streamflow respectively. Raw freshwater inflow values were replaced by 
seasonally adjusted log transformed inflow obtained from the residuals of a GAM model 
fit between season(day of year) and log transformed daily freshwater inflow. Raw nutrient
loads were replaced with flow-adjusted values obtained from the residuals of a
GAM model relating log transformed NO~3~ or TP loads to log transformed daily
inflow. Response residuals from the respective GAM models were 
used as *Q* and *Load* in Eq.\ \ref{eq6} and Eq.\ \ref{eq7}.

@@

Predictive performance of nutrient loads ranged from "satisfactory" to "very good"
based on standardized evaluation metrics of NSE, *r*, and PBIAS 
[@moriasiHydrologicWaterQuality2015] calculated using 5-fold cross validation.
Median goodness-of-fit metrics for NO~3~ models in the Lavaca 
River were 0.34 NSE, 0.70 *r*, and 2.00 PBIAS. Navidad River NO~3~ models 
appeared to perform slightly better with 0.48 NSE and 0.87 *r* but with higher 
bias at 10.90 PBIAS. Generally, TP models performed better than NO~3~ models. 
Median goodness-of-fit metrics for TP in the Lavaca River were 0.81 NSE, 
0.93 *r*, and -7.20 PBIAS. Navidad River TP models had similar performance 
with 0.91 NSE, 0.99 *r*, and -3.30 PBIAS. Density plots of
metrics show similar distribution of values between sites for the 
same parameter, with the exception *r* values for NO~3~ loads where 
Lavaca River had a much larger variance in values compared to the Navidad River
(Fig.\ \ref{fig:fig2}). TP GAMS had higher average NSE and *r* values and 
lower variance in metric values compared to NO~3~.

```{r fig2, warning=FALSE, message=FALSE, fig.width=6.85, fig.asp=0.65, fig.cap="Density plots of goodness-of-fit metrics (NSE, \\textit{r}, and PBIAS) from repeated 5-fold cross validation between predicted nutrient loads from GAM models and measured nutrient loads. Color indicates the tail probability calculated from the empirical cumulative distribution of the goodness-of-fit metrics.", dev='cairo_pdf'}

@@

```{r fig3, error=FALSE, warning=FALSE, message=FALSE, fig.width=6.85, fig.asp=0.7, fig.cap="Aggregated estimated annual and flow-normalized annual NO\\textsubscript{3} and TP loads for USGS-08164000 and USGS-08164525.", dev='cairo_pdf'}

@@

Aggregated across both sites, the mean 
annual NO~3~ load from 2005 through 2020 was `r mean_no3` kg (`r mean_no3_low` kg - 
`r mean_no3_upper` kg, 90% CI). Annual NO~3~ loads ranged from `r min_no3` kg 
in `r min_no3_year` to `r max_no3` kg in `r max_no3_year`. Total annual TP loads 
ranged from `r min_tp` kg in `r min_tp_year` to 
`r max_tp` kg  in `r max_tp_year`. Mean annual TP loading from 2005 through 2020
was `r mean_tp` kg (`r mean_tp_low` kg - `r mean_tp_upper` kg, 90% CI).
On average, the Navidad River accounted for `r no3_mean_prop` of NO~3~ loads and 
`r tp_mean_prop` of TP loads from 2005 through 2020. However, during periods of extreme 
drought the Lavaca River became the primary source of 
nutrient loading in the watershed with the Navidad River only accounting for 
`r min_no3_prop` and `r min_tp_prop` of NO~3~ and TP loads in 2011 
(Fig.\ \ref{fig:fig4}).

```{r fig4, error=FALSE, message=FALSE, warning=FALSE, fig.width=3.3, fig.asp=2, fig.cap="Comparison of delivered annual loads and annual discharge at the Lavaca (USGS-08164000) and Navidad (USGS-08164525) Rivers.", out.width="0.5\\linewidth", dev='cairo_pdf'}

@@

```{r fig5, message=FALSE, warning=FALSE, fig.width=6.85, fig.asp=0.7, fig.cap="Fitted splines (shaded regions indicate 90\\% confidence intervals) from the temporal estuary GAM display the marginal smoothed effect of date on TP (A), NO\\textit{\\textsubscript{X}} (B), chlorophyll-\\textit{a} (C), TKN (D), and DO (E) concentrations at each site in Lavaca Bay.", dev='cairo_pdf'}

@@

```{r fig6, message=FALSE, warning=FALSE, fig.width=6.85, fig.asp=0.7, fig.cap="Fitted splines from estuary GAMs display the marginal smoothed effect of freshwater inflow (controlled for season) on TP (A), NO\\textit{\\textsubscript{X}} (B), chlorophyll-\\textit{a} (C), TKN (D), and DO (E) concentrations at each site in Lavaca Bay.", dev='cairo_pdf'}

@@

```{r fig7, message=FALSE, warning=FALSE, fig.width=6.85, fig.asp=0.44, fig.cap="Fitted splines from the nutrient loading GAMs display the marginal smoothed effect of TP and NO\\textsubscript{3} loads on TP (A) and NO\\textsubscript{\\textit{X}} (B) concentrations at each site in Lavaca Bay.",  dev='cairo_pdf'}

@@

knitr::include_graphics("Schramm-2023-06-PeerJ_files/figure-latex/fig8.png")

@@

The non-linear estuary water quality trends identified in the current
study differed slightly from previously identified trends [@bugica_water_2020]. 
This is not unexpected due to the different time periods, different methodology, and
generally small slopes identified for most of the significant water quality
parameters in prior work. Both DO and cholorophyll-*a* concentrations at all three
Lavaca Bay sites were stable from 2005 through 2020. This is a positive outcome 
in comparison to other Texas estuaries that are facing larger demands for 
freshwater diversions, higher population growth, and more intense agricultural 
production which have resulted in more direct signs of eutrophication
[@wetzWaterQualityDynamics2016; @bugica_water_2020]. Despite the stability of 
DO and cholorophyll-*a*, there are concerning site specific increases in 
NO~*X*~ and TKN concentration over the same time period. These trends are especially
concerning due to the nitrogen limitation identified in many Texas 
estuaries [@gardnerNitrogenFixationDissimilatory2006; 
@houTransformationFateNitrate2012; @doradoUnderstandingInteractionsFreshwater2015; 
@paudelRelationshipSuspendedSolids2019; @wetz_exceptionally_2017] 
and the relatively low ambient concentrations observed in Lavaca Bay.
